---
title: "Allowing for 2 consecutive mismatches in the MH"
output:
  md_document:
    variant: markdown_github
---

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, tidy=TRUE)
```

# Allowing for 2 consecutive mismatches in the MH

MHcut was run twice: allowing for maximum one consecutive mismatch and allowing for 2.
Here both output are compared to quantify how much we gain/lose.

```{r load}
## For data manipulation
library(data.table)
library(dplyr)
library(magrittr)

## For graphs and tables
library(ggplot2)
library(knitr)

## Functions
winsor <- function(x, u=10){
  if(any(x>u)) x[x>u] = u
  x
}
```

We load the coordinate column, rs ID and the MH columns that we want to compare (mhL, mh1L, hom, mhDist).

```{r}
## read.table('../data/mhcut-dbsnp-clinvar-deletion-variants.tsv.gz', nrows=1)
var1 = fread('gunzip -c ../data/mhcut-dbsnp-clinvar-deletion-variants.tsv.gz', select=c(1:4,20:23,25))
var2 = fread('gunzip -c ../data/mhcut-dbsnp-clinvar-deletion-2MM-variants.tsv.gz', select=c(1:4,20:23,25))
```

We then merge the two output by coordinate and rs ID. 
This takes a few minutes, there are 2x43M to merge after all.

```{r}
setkey(var1, chr, start, end, RS, varL)
setkey(var2, chr, start, end, RS, varL)
colnames(var2)[6:9] = paste0(colnames(var2)[6:9], '.2')
var = merge(var1, var2)
var = var[, .(chr, start, end, RS, varL, mh1L, mh1L.2, mhL, mhL.2, hom, hom.2, mhDist, mhDist.2)]
head(var)
```

## Same *mh1L*?

Allowing for consecutive mismatches shouldn't affect the first stretch of exact matches in theory.

```{r}
mean(var$mh1L == var$mh1L.2)
var.m1 = as.data.frame(var[mh1L != mh1L.2,])
```

There are `r nrow(var.m1)` variants with different *mh1L*. Hum...

My only guess is that the selection of the *best* flanks was different.
Allowing for 2 consecutive mismatches produced sometimes a longer MH for the alternative flank pair which was then selected, leading to different MH metrics.
Now I regret not having a column which record which flanks were used (outer-inner or inner-outer).

Maybe I can recompute the score that was used to chose the best flank.
If the score in the 2-mismatches run is always higher, then it fits with the explanation.
The score gives a weight of 2 for the first stretch of exact match and 1 to other matches.

```{r}
var.m1 %>% mutate(score=mh1L + hom*mhL, score.2=mh1L.2 + hom.2*mhL.2) %>% ggplot(aes(x=score, y=score.2)) + geom_point(alpha=.4) + theme_bw() + geom_abline()
```

That fits with the *flank selection* explanation.


## Longer micro-homology?

```{r}
mean(var$mhL.2 > var$mhL)
```

Around 2.2% of the variants have a longer MH when allowing for 2 mismatches.

Graphs with a random subset of the data:

```{r}
var.ss = var[sample.int(nrow(var), 1e5),]
ggplot(var.ss, aes(x=winsor(mhL, 20), y=winsor(mhL.2, 20))) + geom_bin2d(binwidth=c(1,1)) + theme_bw()
ggplot(var.ss, aes(x=mhL.2-mhL)) + geom_histogram() + theme_bw()
var.ss %>% filter(mhL!=mhL.2) %>% ggplot(aes(x=mhL.2-mhL)) + geom_histogram() + theme_bw() + ggtitle('Variants with same mhL removed')
```

Any with shorter MH?

```{r}
sum(var$mhL.2 < var$mhL)
var[mhL.2<mhL,]
```

Just a few and most likely because of the *flank selection*.

## Shorter mhDist?

If the MH length is longer, the distance between MH should be smaller.

Graphs with a random subset of the data:

```{r}
ggplot(var.ss, aes(x=mhDist, y=mhDist.2)) + geom_bin2d() + theme_bw() + geom_abline()
ggplot(var.ss, aes(x=mhDist.2-mhDist)) + geom_histogram() + theme_bw()
var.ss %>% filter(mhL!=mhL.2) %>% ggplot(aes(x=mhDist.2-mhDist)) + geom_histogram() + theme_bw() + ggtitle('Variants with same mhL removed')
```

As expected most of the variants that were further extended have a shorter distance between their MH.


## Lower homology?

Graphs with a random subset of the data:

```{r}
ggplot(var.ss, aes(x=hom, y=hom.2)) + geom_bin2d() + theme_bw() + geom_abline()
ggplot(var.ss, aes(x=hom.2-hom)) + geom_histogram() + theme_bw()
var.ss %>% filter(mhL!=mhL.2) %>% ggplot(aes(x=hom.2-hom)) + geom_histogram() + theme_bw() + ggtitle('Variants with same mhL removed')
```

As expected most of the variants that were further extended have a lower homology.


## Effect on a typical filtering example

Let's say we want variants with a MH starting with at least 3 exact matches (**mh1L>2**) and a maximum MHdist of 5 (**mhDist<6**).
How many variants do we "gain" by allowing 2 mismatches in the MH?

```{r}
nb1 = nrow(var[mh1L>2 & mhDist<6,])
nb2 = nrow(var[mh1L.2>2 & mhDist.2<6,])
nb2-nb1
nb2/nb1
```

We get `r nb2-nb1` more variants, a `r round((nb2/nb1-1)*100, 2)`% increase.
