---
title: "Chromosome view of the variants"
output:
  md_document:
    variant: markdown_github
---

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, tidy=TRUE)
```

# Chromosome view of the variant density

```{r load}
## For data manipulation
library(data.table)
library(dplyr)
library(magrittr)

## For graphs and tables
library(ggplot2)
library(knitr)

## Functions
winsor <- function(x, u=10){
  if(any(x>u)) x[x>u] = u
  x
}
```

We load the coordinate columnd and the MH columns that we want to use for filtering (e.g. mhL and mh1L).

```{r}
## read.table('../data/mhcut-dbsnp-clinvar-deletion-variants.tsv.gz', nrows=1)
var = fread('gunzip -c ../data/mhcut-dbsnp-clinvar-deletion-variants.tsv.gz', select=c(1:3,21:22))
setkey(var, chr)
```

## Variant density

For example, density of variants with `mh1L>2` in region of 1 Mbp.

```{r}
## source("https://bioconductor.org/biocLite.R")
## biocLite("BSgenome.Hsapiens.UCSC.hg38")
library(BSgenome.Hsapiens.UCSC.hg38)
library(GenomicRanges)
```

First let's prepare 1 Mbp regions in the human genome.

```{r cache=TRUE}
bin.gr = tileGenome(seqlengths(BSgenome.Hsapiens.UCSC.hg38), tilewidth=1e6)
bin.gr = do.call(c, bin.gr)
```

Now count how many variants overlap each 1 Mbp region.

```{r cache=TRUE}
var.filt = var[mh1L>2,]
## Separately per chromosome to reduce the memory footprint
gr.l = lapply(paste0('chr',c(1:22, 'X', 'Y')), function(chrom){
  var.gr = makeGRangesFromDataFrame(var.filt[chrom,])
  bin.chr = bin.gr[which(seqnames(bin.gr)==chrom)]
  bin.chr$nb.var = countOverlaps(bin.chr, var.gr)
  bin.chr
})

## Merge the chromosomes
gr = do.call(c, gr.l)
head(gr)
```

On average `r round(mean(gr$nb.var))` variants per Mbp.

## IdeoViz package


```{r}
## source("https://bioconductor.org/biocLite.R")
## biocLite("IdeoViz")
library(IdeoViz)

ideo <- getIdeo("hg38")

plotOnIdeo(chrom = paste0('chr', 1:11),
           ideoTable = ideo,
           values_GR = gr,
           value_cols = colnames(mcols(gr)),
           col = 'indianred2',
           addScale = F,
           val_range=c(0, max(gr$nb.var)),
           ## plotType='lines',
           plotType='rect',
           ## plot_title = paste("Differrence between", gsub(".coverage","",fileA), gsub(".coverage","",fileB), "bin",windows_size),
           cex.axis = 0.3,
           chromName_cex = .6,
           vertical = T)

plotOnIdeo(chrom = paste0('chr', c(12:22, 'X', 'Y')),
           ideoTable = ideo,
           values_GR = gr,
           value_cols = colnames(mcols(gr)),
           col = 'indianred2',
           addScale = F,
           val_range=c(0, max(gr$nb.var)),
           ## plotType='lines',
           plotType='rect',
           ## plot_title = paste("Differrence between", gsub(".coverage","",fileA), gsub(".coverage","",fileB), "bin",windows_size),
           cex.axis = 0.8,
           chromName_cex = 0.6,
           vertical = T)
```


## chromPlot package

```{r}
## source("https://bioconductor.org/biocLite.R")
## biocLite("chromPlot")
library(chromPlot)

## (Down)Load band information from UCSC
if(!file.exists('../data/cytoBandIdeo-hg38.txt.gz')){
  download.file('http://hgdownload.soe.ucsc.edu/goldenPath/hg38/database/cytoBandIdeo.txt.gz', '../data/cytoBandIdeo-hg38.txt.gz')
}
cyto = read.table('../data/cytoBandIdeo-hg38.txt.gz', sep='\t')
colnames(cyto) = c('Chrom','Start','End','Name','gieStain')
cyto = subset(cyto, Chrom %in% paste0('chr', c(1:22, 'X','Y')))

## (Down)Load gap information from UCSC
if(!file.exists('../data/gap-hg38.txt.gz')){
  download.file('http://hgdownload.soe.ucsc.edu/goldenPath/hg38/database/gap.txt.gz', '../data/gap-hg38.txt.gz')
}
gap = read.table('../data/gap-hg38.txt.gz', sep='\t')
gap = gap[,c(2:4,8)]
colnames(gap) = c('Chrom','Start','End','Name')
gap = subset(gap, Chrom %in% paste0('chr', c(1:22, 'X','Y')))

## Point graph
chromPlot(gaps=gap, bands=cyto,
          stat=gr,
          statCol='nb.var',
          statName='variants')

## Line graph
chromPlot(gaps=gap, bands=cyto,
          stat=gr,
          statCol='nb.var',
          statTyp='l',
          statName='variants')
```
